The markings of the trade: exploring the patterns of olive oil production in Roman Baetica

### 2020-04-28 Paper/jaccardbaetica.tex
new changes intro xavi

Material culture is one of the most frequent indicators of trade in the archaeological record. In archaeology, they allow us to highlight a part of the mechanism of production and distribution of goods along the Mediterranean \citep{bevan_mediterranean_2014}. Particularly, the spread of these factors had an important impact during the Roman Age, when the progressive exploitation of communication networks allowed a major interaction between communities \citep{orengo_seeds_2016}\xavi{aquí faltan citas de obras más generales; esto es un ejemplo pero muy específico noÇ?}. Given this fact, an important mechanism of production control under the Roman government was created spreading by different areas attending to the richness of different places. Roman provinces such as \textit{Baetica} (currently Andalusia, south Spain) became important production centres of olive oil during the Roman Empire. Olive oil was considered the liquid gold since it was used in different aspect of daily life for cooking, hygiene or lighting  \citep{mattingly_d.j._oil_1988}. As a consequence, this product was distributed in a large number of amphorae along the province, mostly to supply the Roman Army and Italy \citep{blazquez_exportacion_1980}.

Despite this issue has been widely discussed over many decades, the understanding of the olive oil production processes is still under debate, mostly due to the lack of written records. The application of different quantitative approaches combined with the archaeological evidence has allowed us partly to deal with the complexity of the Roman production \citep{orengo_seeds_2016,
brughmans_roman_2016,coto-sarmiento_identifying_2018,bayesian_2018}.

>>>>

%spread of goods and ideas had a enormous impact during the Roman Empire. The progressive exploitation of communication networks allowed a major interaction between communities
%This can be seen by the fact that material culture is found in different regions along the Mediterranean as a frequent indicator of trade exchange in the archaeological record.

The role of the Mediterranean sea for the trade had an enormous impact during the Roman Empire. The spread of thousands of different goods along maritime shipping allowed a progressive exploitation of communication networks and a major interaction between communities \citep{rodriguez_baetican_1998, 
temin_market_2001,
bevan_mediterranean_2014}. The exchange of different products between regions can be seen in both archaeological and written records as a frequent indicator of Roman trade.

%Material culture is one of the most frequent indicators of trade in the archaeological record. In archaeology, they allow us to highlight a part of the mechanism of production and distribution of goods along the Mediterranean \citep{bevan_mediterranean_2014}. Particularly, the spread of these factors had an important impact during the Roman Age, when the progressive exploitation of communication networks allowed a major interaction between communities \citep{orengo_seeds_2016}\xavi{aquí faltan citas de obras más generales; esto es un ejemplo pero muy específico no?}.

Given this fact, an important mechanism of production control under the Roman government was created spreading by different areas attending to the richness of different places. Roman provinces such as \textit{Baetica} (currently Andalusia, south Spain) became important production centres of olive oil during the Roman Empire. Olive oil was considered the liquid gold since it was used in different aspect of daily life for cooking, hygiene or lighting  \citep{mattingly_d.j._oil_1988}. As a consequence, this product was distributed in a large number of amphorae along the provinces, mostly to supply the Roman Army and Italy \citep{blazquez_exportacion_1980}.

The understanding of the olive oil production processes has been widely discussed over many decades\citep{rodriguez_economioleicola_1977, Chic_hispania_1997,millet_anforas_1998}. Despite this topic remains under debate: what patterns were followed for the distribution of olive oil to the different provinces?
Did each province follow a different pattern for its distribution? Nor the lack of written records has made possible to detect any indication of patterns in the olive oil market.

Advances in the research of the Roman world have currently led an environment with more diverse commercial dynamics \citep{duncan1982economy,
temin_economy_2006,
quantifyingwilson2009}. The application of different quantitative approaches combined with the archaeological evidence has allowed us partly to deal with the complexity of the Roman production \citep{brughmans_roman_2016,
orengo_seeds_2016,bayesian_2018,
coto-sarmiento_identifying_2018,
rubio-campillo_ecology_2018}.
